The Changing Landscape of Cystic Fibrosis Related Diabetes Risk Factors Between 2003-2018

### 2025-06-15 Christine Chan/CF Registry/WNAR/wnar_presentation.qmd
added a few WNAR slides

@@

$$
\lambda(t)=\lambda_0(t)e^{\beta(t)X}
$$

@@

HOW MUCH MATH DO I NEED HERE?

-   Coefficients vs. covariates?
-   Internal vs. external covariates?

# An Approach That Did Not Work

-

@@

```{r}
#| fig-width: 12
#| fig-height: 9
comb_plot <- ggcoef_compare(
  list("2003 - 2008" = era1, "2009 - 2013" = era2, "2014 - 2018" = era3),
  add_reference_rows = T, exponentiate = T, shape_guide = F,colour=F,
  type = "faceted"
)
comb_plot
```

>>>>

# What do we do when the proportional hazards assumption is violated?

## Two good options

1. Joint modeling of longitudinal and survival data.[@hendersonJointModellingLongitudinal2000]
  - The gold standard, do this if you can.
  - Complex, requires somewhat specialized knowledge.
  - Ask Elizabeth for help with this.

2. Time-dependent Cox models
  - More appropriate for external covariates (measures that vary independent of the failure time).
  - Or baseline covariates where the effect changes over time.[@zhangTimevaryingCovariatesCoefficients2018]

@@

$$
\lambda(t)=\lambda_0(t)exp(\beta(t)X)
$$

@@

- Another way to write this model:

$$
\lambda(t)=\lambda_0(t)exp(g(\beta,t)X)
$$

- Where $g(\beta,t)$ is a function of time that we can define.[@zhangTimevaryingCovariatesCoefficients2018]

## Time as a Step Function

```{r}
data(cancer)
vet2 <- survSplit(Surv(time, status) ~ .,
  data = cancer, cut = c(90, 180), episode = "tgroup", id = "id"
)
vfit2 <- coxph(Surv(tstart, time, status) ~ age + sex + ph.ecog +
  pat.karno:strata(tgroup), data = vet2)
vfit2
```

# Applying This to Our CFRD Data

## The Code

```{r}
annual_split <- survSplit(Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2) ~ .,
  data = annual, cut = c(2192, 4018), episode = "tgroup"
)
cox_res2 <- coxph(
  Surv(time = tstart, time2 = tstop, event) ~
    age_group:strata(tgroup) + Gender + Race + Hispanicrace + bmi_cat + 
    MutClass:strata(tgroup) + A_IsOnEnzymes:strata(tgroup) + 
    tube_feeding:strata(tgroup) + liver_disease:strata(tgroup) + 
    ppFEV1_cat:strata(tgroup) + NumPulmExacerbation:strata(tgroup) + 
    A_corticosteroids1:strata(tgroup),
  data = annual_split, cluster = eDWID
)
```

## The Results

```{r}
cox_res2 %>%
  tbl_regression(exponentiate = T)
```

## The Interpretation

![](homer.jpg)

@@

```{r}
#| fig-width: 12
#| fig-height: 9
comb_plot <- ggcoef_compare(
  list("2003 - 2008" = era1, "2009 - 2013" = era2, "2014 - 2018" = era3),
  add_reference_rows = T, exponentiate = T, shape_guide = F, colour = F,
  type = "faceted"
)
comb_plot
```
